---
title: "Higher Education's Fall 2020 Plans"
author: "Sarah Gillespie"
date: "6/18/2020"
output:
  html_document: default
---

```{r include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
# setting up the file
require(knitr)
opts_chunk$set(
  tidy=TRUE,                     # display code as typed
  size="small",                   # slightly smaller font for code
  tidy.opts=list(width.cutoff=65), # wrap text and long comments
  fig.width=7, fig.height=5           #figure size
)

knitr::opts_chunk$set(echo = TRUE)

```

```{r include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
require(tidyverse)
require(mosaic)
require(mosaicData)
require(lmtest)
library(ggthemes)
library(formatR)
library(gmodels) 
library(Hmisc)
```

I attend a small college and am apprehensively waiting to know my fall plans. I wanted to see if there was a natural consensus among colleges with similar attributes to my own school by looking at current Fall 2020 plans broken down by geographic location, acceptance rate, average SAT score, percent of international students, and median household income. Unfortunately for me, the most common announcement among those other schools is that they’re "waiting to make the decision" or "considering a range of scenarios." This is most prominent for schools with a low acceptance rate or a high percentage of international students.

School plans were sourced from [The Chronicle of Higher Education’s frequently updated list](https://www.chronicle.com/article/Here-s-a-List-of-Colleges-/248626?utm_source=at&utm_medium=en&utm_source=Iterable&utm_medium=email&utm_campaign=campaign_1209367&cid=at&source=ams&sourceId=378492) tracking nearly a thousand school’s Fall 2020 plans. The list did not specify how each school was selected and there is an option to submit untracked institutions to the list, which means the large sample size is not necessarily randomly selected from all the higher education institutions in the United States. The sample size does include many well-known colleges, including all 23 Cal State campuses, and many Massachusetts schools (MIT, Smith, Amherst, Wellesley, Hampshire, Boston University, Boston College, Brandeis, Northeastern) but also many lesser-known ones across the U.S., such as Point Loma Nazarene University and the University of South Alabama, and community colleges.

Last time I downloaded the data: June 18

[My Github project with code](https://github.com/SarahGillespie/covidPlansForFall)


```{r include=FALSE, warning=FALSE, message=FALSE, error=FALSE}

#read in the article data
fall2020dataCoHEJun18 <- read_csv("fall2020dataCoHEJun18.csv")

#remove any doubles in the article's data
fall2020dataCoHEJun18 <- unique(fall2020dataCoHEJun18)

#separate anything with a dash, like California State University-Fresno with the specific campus being campus1
fall2020dataCoHEJun18 <- fall2020dataCoHEJun18 %>%
  separate(Institution, c("mainname", "campus1"), " — ")

#fall2020dataCoHEJun18 <- fall2020dataCoHEJun18[!duplicated(fall2020dataCoHEJun18$mainname), ]

#read in government data
Most_Recent_Cohorts_SMALL <- read_csv("Most-Recent-Cohorts-SMALL.csv")

#rename the school name to be institution
Most_Recent_Cohorts_SMALL <- rename(Most_Recent_Cohorts_SMALL, Institution = INSTNM)

#separate anything with a dash, like California State University-Fresno with the specific campus being campus2 that is then removed
Most_Recent_Cohorts_SMALL <- Most_Recent_Cohorts_SMALL %>%
  separate(Institution, c("mainname", "campus2"), "-") %>%
  select(-campus2)

Most_Recent_Cohorts_SMALL <- Most_Recent_Cohorts_SMALL[!duplicated(Most_Recent_Cohorts_SMALL$mainname), ]
 

KaggleRegions <- read_csv("KaggleRegions_datasets_31195_39789_states.csv")

KaggleRegions <- rename(KaggleRegions, StateL = State)

KaggleRegions <- rename(KaggleRegions, "State" = "State Code")

bigList<-merge(x=fall2020dataCoHEJun18,y=Most_Recent_Cohorts_SMALL,by="mainname",all.x=TRUE)

bigList<-merge(x=bigList,y=KaggleRegions,by="State",all.x=TRUE)

bigList<-bigList[!is.na(bigList$STABBR),]



List <- bigList %>%
  separate(Category, c("linkish", "Call1"), ">") %>%
  separate(Call1, c("Call", "extra"), "<") %>%
  select(-linkish, -extra, -UNITID, -OPEID, -OPEID6, -DISTANCEONLY, -RELAFFIL, -HCM2, -CCSIZSET, -CCUGPROF, -CCBASIC)

#makes decision all lowercase
List[[5]] <- tolower(List[[5]])

#2016 election results
X2016results <- read_csv("2016results.csv")

X2016results <- rename(X2016results, StateL = State)


#join election results to each school by state

List<-merge(x=List,y=X2016results,by="StateL",all.x=TRUE)
```

```{r warning=FALSE, message=FALSE, error=FALSE}

#ALL STATES ALL SCHOOLS INCLUDING SOME GRAD SCHOOLS

#by state

# ggplot(List, aes(y = State)) +
#  geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))

#region

ggplot(List, aes(y = Region)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE)) +
  labs(title="Colleges fall 2020 call by Region", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab("Region of the U.s.")

#division

ggplot(List, aes(y = Division)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE)) +
  labs(title="Colleges fall 2020 call by Division", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab(" ")

#public vs. private

ggplot(List, aes(y = Control)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))  +
  labs(title="Public vs. Private school fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab(" ")

#all colleges choices

ggplot(List, aes(y = Call)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))  +
  labs(title="All schools' fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab(" ")

#women only
WomenOnly <- List %>%
  filter(WOMENONLY == 1)

ggplot(WomenOnly, aes(y = Call)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE)) +
  labs(title="Historically Women's Colleges' fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab(" ")


#Massachusetts only

MassOnly <- List %>%
  filter(State == "MA")

ggplot(MassOnly, aes(y = Call)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE)) +
  labs(title="Massachusetts schools' fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few() +
  xlab("Number of schools") +
  ylab(" ")

```

```{r warning=FALSE, message=FALSE, error=FALSE}

#scatterplots


List$ADM_RATE <- as.numeric(as.character(List$ADM_RATE))
List$MEDIAN_HH_INC <- as.numeric(as.character(List$MEDIAN_HH_INC))
List$POVERTY_RATE <- as.numeric(as.character(List$POVERTY_RATE))
List$UGDS <- as.numeric(as.character(List$UGDS))
List$percR <- as.numeric(as.character(List$percR))
List$ADM_RATE <- as.numeric(as.character(List$ADM_RATE))
List$PCT_BORN_US <- as.numeric(as.character(List$PCT_BORN_US))


#admin rate and median HH income
ggplot(List, aes(x = ADM_RATE, y = MEDIAN_HH_INC, color = Call)) +
  geom_point() +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Median Household Income, acceptance rate, and fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few()  +
  xlab("Acceptance Rate") +
  ylab("Median Household Income")

#admin rate and poverty rate

# ggplot(List, aes(x = ADM_RATE, y = POVERTY_RATE, color = Call)) +
#   geom_point() +
#   scale_y_continuous(labels = scales::comma)

#admin rate and number of undergrad students

# ggplot(List, aes(x = ADM_RATE, y = UGDS, color = Call)) +
#   geom_point() +
#   scale_y_continuous(trans = 'log10')

#admin rate and percentage of students born inside the United States
ggplot(List, aes(x = ADM_RATE, y = PCT_BORN_US, color = Call)) +
  geom_point() +
  scale_y_continuous(trans = 'log10') +
  ylim(65,100) +
  labs(title="Percent of students born in the U.S., acceptance rate, and fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few()  +
  xlab("Acceptance Rate") +
  ylab("Percentage of students born in the U.S.")

```



```{r warning=FALSE, message=FALSE, error=FALSE}

# selective schools: I define this as schools with an acceptance rate of <25%.

SelectiveSchool <- List

SelectiveSchool <- data.frame(lapply(SelectiveSchool, function(x) {
                  gsub("NULL", "", x)}))

SelectiveSchool$percR <- as.numeric(as.character(SelectiveSchool$percR))
SelectiveSchool$ADM_RATE <- as.numeric(as.character(SelectiveSchool$ADM_RATE))
SelectiveSchool$SAT_AVG <- as.numeric(as.character(SelectiveSchool$SAT_AVG))

SelectiveSchool <- SelectiveSchool %>%
  filter(ADM_RATE < 0.25)


#by state

# ggplot(SelectiveSchool, aes(y = State)) +
#  geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))

#region

ggplot(SelectiveSchool, aes(y = Region)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))+
  labs(title="Selective schools' fall 2020 call by region", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few()  +
  xlab("Number of schools") +
  ylab(" ")

#division

# ggplot(SelectiveSchool, aes(y = Division)) +
#  geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))

#public vs. private

# ggplot(SelectiveSchool, aes(y = Control)) +
#  geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))

#all colleges choices

# ggplot(SelectiveSchool, aes(y = Call)) +
#  geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE))

#massachusetts only

MassOnly <- SelectiveSchool %>%
  filter(State == "MA")

ggplot(MassOnly, aes(y = Call)) +
 geom_bar(aes(fill = Call), position = position_stack(reverse = FALSE)) +
  labs(title="Massachusetts selective schools' fall 2020 call", caption = "Data from The Chronicle of Higher Education on 6/18/2020") +
  theme_few()  +
  xlab("Number of schools") +
  ylab(" ")

#admin rate and state's political lean

# ggplot(SelectiveSchool, aes(x = ADM_RATE, y = percR, color = Call)) +
#   geom_point()


```


I joined four data sets when making the data visualizations. 
I began with The Chronicle of Higher Education article’s data. I opened it in Microsoft Excel to change the formatting to account for each campus for schools with multiple campuses, such as the California State University system.

I used the [United States Department of Education’s most recent institution-level data](https://collegescorecard.ed.gov/data/) and joined it with the Chronicle of Higher Education data by institution name. I first brought the Department of Education data into Excel and removed unnecessary columns because the downloaded csv was too large for my Rstudio desktop version to open.

I joined a [Kaggle spreadsheet created by Omer Shacham](https://www.kaggle.com/omer2040/usa-states-to-region) to categorize each school’s state into regions and divisions of the United States.
 
I joined the [World Population Review 2016 Election Results By State](https://worldpopulationreview.com/states/2016-election-results-by-state/) spreadsheet for each state’s presidential election results.


Later, I hope to break it down by what the school is ranked in education and healthcare costs. I am also interested in including multiple linear regression once more schools release their fall 2020 plans.

